Stability concerns in random matrices

### 2018-03-17 random_matrix_stability.Rmd
Re-orient notes

@@

The above is a draft introductory paragraph or abstract. I believe that we could get [an article on the above](https://www.nature.com/natecolevol/about/content) clearly in the space of 3500, or maybe even 1500 words. As far as I've been able to find, such an article would be novel and interesting (and mathematically correct).

Background on the key question
--------------------------------------------------------------------------------

@@

Additionally, I want to look at random ecological matrices from the context of varying generation time, as partly inspired by the use of $\varepsilon$ by @Patel2018 and partly by @Jirsa2004 to separate generation time of different species in a community. Species generation time varies by orders of magnitude within communities, potentially affecting the rate at which the density of one species affects the population growth rate of the other. Yet I've not been able to find anything on this in the literature in a random matrix context. I want to **determine if the effects of varying generation time among community members is stabilising or destabilising in an ecological system**. Hence, the goals here are twofold.

1. Demonstrate the inherent destabilising effect of rapid evolution, which can also be interpreted as plasticity [@Patel2018], on ecological communities.

2. Determine if variance in generation time stabilises or destabilises random matrices.

I'm not sure if either would make a publication, or perhaps just a blog post, and whether they would need to be completely separated. I will present one a time.

@@

The effect of variation in generation time on ecological stability
--------------------------------------------------------------------------------

Here I will only consider $\boldsymbol{A}$ and focus on the stability of ecological interactions given variation in generation time. I will use the same technique of @Patel2018 for varying ecological and evolutionary time scales, but a bit differently. I assume that the response of a particular species' growth rate (rows in $\boldsymbol{A}$) is consistently affected by a value $\gamma_{i}$ (I am using $\gamma_{i}$ instead of $\varepsilon_{i}$ to avoid confusing generation time variance with ecological versus evolutionary rate difference). Higher values of $\gamma_{i}$ correspond to shorter absolute generation times and lower values correspond to longer generation times. The matrix below $\boldsymbol{A}$ for two species $i$ and $j$ with generation times $\gamma_{i}$ and $\gamma_{j}$, respectively.

>>>>

The below is a draft introductory paragraph or abstract. I believe that we could get [an article on the above](https://www.nature.com/natecolevol/about/content) clearly in the space of 3500, or maybe even 1500 words. As far as I've been able to find, such an article would be novel and interesting (and mathematically correct). These notes are divided into two parts at the moment.

1. [The effect of variation in generation time on community stability.](#gen_time_stability)
2. [The effect of increasing numbers of evolutionary traits on eco-evolutionary stability.](#comment_on_Patel)

I suspect that part 2 will need to branch off soon, but I don't want to forget about it.

@@

<a name="gen_time_stability">The effect of variation in generation time on community stability</a>
--------------------------------------------------------------------------------

Species generation time varies by orders of magnitude within communities, potentially affecting the rate at which the density of one species affects the population growth rate of the other. Yet I've not been able to find anything on this in the literature in a random matrix context. I want to **determine if the effects of varying generation time among community members is stabilising or destabilising in an ecological system**. This work is partly inspired by the use of $\varepsilon$ by @Patel2018 and partly by @Jirsa2004 to separate generation time of different species in a community.

Here I will consider a Jacobian matrix of interacting species $\boldsymbol{A}$ and focus on the stability of ecological interactions given variation in generation time. I will use the same technique of @Patel2018 for varying ecological and evolutionary time scales, but a bit differently. I assume that the response of a particular species' growth rate (rows in $\boldsymbol{A}$) is consistently affected by a value $\gamma_{i}$ (I am using $\gamma_{i}$ instead of $\varepsilon_{i}$ to avoid confusing generation time variance with ecological versus evolutionary rate difference). Higher values of $\gamma_{i}$ correspond to shorter absolute generation times and lower values correspond to longer generation times. The matrix below $\boldsymbol{A}$ for two species $i$ and $j$ with generation times $\gamma_{i}$ and $\gamma_{j}$, respectively.

<a name="comment_on_Patel">The effect of increasing numbers of evolutionary traits on eco-evolutionary stability</a>
--------------------------------------------------------------------------------

@@

Hence, the goal here is to demonstrate the inherent destabilising effect of rapid evolution, which can also be interpreted as plasticity [@Patel2018], on ecological communities.

@@
